### 2019-02-27 sections/introduction.tex
push fixmes in margin for last proofread

\fixme{Alejandro: The introduction seems to be more about why Task-Oriented Programming
is nice and useful. Personally, I would like it to tell me another
story: given that some people find TOP useful (as shown by the many
examples), why should be care about formalizing it and which challenges
does it pose? I think this is what people from ICFP would care about more.}

Many applications these days are developed to support workflows in institutions and businesses.
Take for example expense declarations, order processing, and emergency management.
Some of these workflows occur on the boundary between organisations and customers,
like flight bookings or tax returns.
What they all have in common is
that they need to interact with different people (end users, tax officers, customers, etc.)
and they use information from multiple sources (input forms, databases, sensors, etc.).

@@

\subsection{Task-oriented programming}
\fixme{Alejandro: Not entirely convincing}%Nico: hij heeft het hier volgens mij over de titel van de sectie en het woord aims in de eerste zin. Misschien moeten we een hardere claim doen hier?
% Task-Oriented Programming (\TOP) is a programming paradigm aimed at writing interactive multi-user applications in a declarative way \cite{conf/ppdp/PlasmeijerLMAK12}.
Task-oriented programming (\TOP) is a programming paradigm which aims to be the sweet spot between faithful modelling workflows
and rapid prototyping of multi-user web applications supporting these workflows \cite{conf/ppdp/PlasmeijerLMAK12}.
% Task-oriented programming (\TOP) is a programming paradigm to support these ways of working.
\TOP focusses on modelling collaboration patterns.
This gives rise to user's need to interact and share information.
Next to that, \TOP automatically provides solutions to common development jobs like designing \GUI\ s, connecting to databases, and communicating between servers and clients.

@@

Users can work together in a number of ways, and this is reflected in \TOP by task compositions.
There is sequential composition, parallel composition, and choice.
Users need to communicate in order to engage in these forms of collaboration.
This is reflected in \TOP by three kinds of communication mechanisms.
There is data flow \emph{alongside} control flow, where the result of a task is passed onto the next.
There is data flow \emph{across} control flow, where information is shared between multiple tasks.
Finally, there is communication with the \emph{outside} world, where information is entered into the system via input events.
%\fixme{Should this be mentioned earlier?}
The end points where the outside world interacts with \TOP applications are called editors.
In generated applications, editors can take many forms, like input fields, selection boxes, or map widgets.

>>>>

\todo{Alejandro: The introduction seems to be more about why Task-Oriented Programming
is nice and useful. Personally, I would like it to tell me another
story: given that some people find TOP useful (as shown by the many
examples), why should be care about formalizing it and which challenges
does it pose? I think this is what people from ICFP would care about more.}
Many applications these days are developed to support workflows in institutions and businesses.
Take for example expense declarations, order processing, and emergency management.
Some of these workflows occur on the boundary between organisations and customers,
like flight bookings or tax returns.
What they all have in common is
that they need to interact with different people (end users, tax officers, customers, etc.)
and they use information from multiple sources (input forms, databases, sensors, etc.).

@@

\subsection{Task-oriented programming}
\todo{Alejandro: Not entirely convincing}%Nico: hij heeft het hier volgens mij over de titel van de sectie en het woord aims in de eerste zin. Misschien moeten we een hardere claim doen hier?
% Task-Oriented Programming (\TOP) is a programming paradigm aimed at writing interactive multi-user applications in a declarative way \cite{conf/ppdp/PlasmeijerLMAK12}.
Task-oriented programming (\TOP) is a programming paradigm which aims to be the sweet spot between faithful modelling workflows
and rapid prototyping of multi-user web applications supporting these workflows \cite{conf/ppdp/PlasmeijerLMAK12}.
% Task-oriented programming (\TOP) is a programming paradigm to support these ways of working.
\TOP focusses on modelling collaboration patterns.
This gives rise to user's need to interact and share information.
Next to that, \TOP automatically provides solutions to common development jobs like designing \GUI\ s, connecting to databases, and communicating between servers and clients.

@@

Users can work together in a number of ways, and this is reflected in \TOP by task compositions.
There is sequential composition, parallel composition, and choice.
Users need to communicate in order to engage in these forms of collaboration.
This is reflected in \TOP by three kinds of communication mechanisms.
There is data flow \emph{alongside} control flow, where the result of a task is passed onto the next.
There is data flow \emph{across} control flow, where information is shared between multiple tasks.
Finally, there is communication with the \emph{outside} world, where information is entered into the system via input events.
\todo{Should this be mentioned earlier?}
The end points where the outside world interacts with \TOP applications are called editors.
In generated applications, editors can take many forms, like input fields, selection boxes, or map widgets.

### 2019-02-27 sections/intuition.tex
call it interleaving concurrency to make sure what we mean
make notes about internal discussions
use cleveref consistently
push fixmes in margin for last proofread

The way the combinators are defined matches real life closely.
When we want to have breakfast, we have to complete several other tasks first before we can do so.
We decide what we want to have and then prepare it.
We can prepare the different items we have for breakfast in parallel, but not at the same time.
For example, it is impossible to scramble eggs, and put on the kettle for tea simultaneously.
Instead, what is meant by parallel is that the order in which we do tasks and the smaller tasks that they are composed of, does not matter.
\fixme{Alejandro has placed a line and exclamation under this sentence. I'm not sure what to do here.}
Then finally, only when every item we want to have for breakfast is ready, can we sit down and enjoy it.
\end{example}

@@

In \TOPHAT, editors are the basic method of communication with the environment.
There are different editors, denoted by different box symbols.
An editor of the form $\Edit v$, where $v$ is of type $\tau$, reacts to two kinds of events:
Change events $v$ with a value of type $\tau$ and clear events $\Empty$ which empty the editor.
This results in a typed empty editor $\Enter \tau$, which is annotated with the same type.
\fixme{Alejandro: unclear} %Nico: He has marked the above paragraph as being unclear.

@@

\fixme{Alejandro: What is a value? What is a type?}

@@

Several observations can be made about tasks.
One of those is determining the value of a task $\Value(t)$.
% \fixme{Should we change the type of Value to Maybe v? Both johan and alejandro mention that it is not completely clear to them.}
% I don't think that we should use Maybe. We should use the term partial function and bottom. -- mkl
Not all tasks have a value, for example the empty editor $\Enter \tau$, which makes $\Value$ a partial function.
For example, $\Value(\Edit 5) = 5$, $\Value(\Enter \Int) = \bot$.

@@

Let $t_1$ and $t_2$ be tasks.
The parallel combination $t_1 \And t_2$ stands for two independent tasks carried out at the same time.
Therefore $\And$ introduces concurrency. \fixme{Alejandro: But before you said it was interleaving!}%nico: en later doen we dit ook.
For the system it does not matter if the tasks are executed by two people actually in parallel, or by one person who switches between the tasks.
The inputs sent to the component tasks are interleaved into a serial stream, which is sent to the parallel combinator.
We assume that such a serialization is always possible.
The tasks are truly independent of each other, all interleavings are permitted.
The environment must prefix events to $t_1$ and $t_2$ by $\First$ (first) and $\Second$ (second) respectively.
This unambiguously renames the inputs, removing any source of nondeterminism.

@@

% \lstset{emph={text}}
% \begin{example}[Abstract submission]
% \label{exm:abstract}
%
% The task of writing an abstract for a conference is never done.
% Let \TS{abstract} be a reference to a string.
%
% % \begin{TASK}
% %   let submit : String -> Task Unit =
% %     \text. delay 30 (sendTo "icfp19@sigplan.org" text)
% %   let write : Task Unit =
% %     edit "Abstract: ..." >>= \abstract. if 500 <= words abstract <= 700 then submit abstract else fail
% % \end{TASK}
%
% \begin{TASK}
%   let submit : Ref String -> Task String = \text. sendTo "icfp19@sigplan.org" text in
%   update abstract <|> delay 30 (submit abstract)
% \end{TASK}
%
% The \TS{delay} task is specified in Example~\ref{exm:delay} in Section~\ref{sec:language}.
%
% \end{example}

>>>>

The way the combinators are defined matches real life closely.
When we want to have breakfast, we have to complete several other tasks first before we can do so.
We decide what we want to have and then prepare it.
We can prepare the different items we have for breakfast in parallel, but not at the same time.
For example, it is impossible to scramble eggs, and put on the kettle for tea simultaneously.
Instead, what is meant by parallel is that the order in which we do tasks and the smaller tasks that they are composed of, does not matter.
\todo{Alejandro has placed a line and exclamation under this sentence. I'm not sure what to do here.}
Then finally, only when every item we want to have for breakfast is ready, can we sit down and enjoy it.
\end{example}

@@

In \TOPHAT, editors are the basic method of communication with the environment.
There are different editors, denoted by different box symbols.
An editor of the form $\Edit v$, where $v$ is of type $\tau$, reacts to two kinds of events:
Change events $v$ with a value of type $\tau$ and clear events $\Empty$ which empty the editor.
This results in a typed empty editor $\Enter \tau$, which is annotated with the same type.
\todo{Alejandro: unclear} %Nico: He has marked the above paragraph as being unclear.

@@

\todo{Alejandro: What is a value? What is a type?}

@@

Several observations can be made about tasks.
One of those is determining the value of a task $\Value(t)$.
%NOTE: Should we change the type of Value to Maybe v? Both johan and alejandro mention that it is not completely clear to them. --NN
%NOTE: I don't think that we should use Maybe. We should use the term partial function and bottom. --MKL
Not all tasks have a value, for example the empty editor $\Enter \tau$, which makes $\Value$ a partial function.
For example, $\Value(\Edit 5) = 5$, $\Value(\Enter \Int) = \bot$.

@@

Let $t_1$ and $t_2$ be tasks.
The parallel combination $t_1 \And t_2$ stands for two independent tasks carried out at the same time.
This operator introduces interleaving concurrency.
For the system it does not matter if the tasks are executed by two people actually in parallel, or by one person who switches between the tasks.
The inputs sent to the component tasks are interleaved into a serial stream, which is sent to the parallel combinator.
We assume that such a serialization is always possible.
The tasks are truly independent of each other, all interleavings are permitted.
The environment must prefix events to $t_1$ and $t_2$ by $\First$ (first) and $\Second$ (second) respectively.
This unambiguously renames the inputs, removing any source of nondeterminism.

@@

% \lstset{emph={text}}
% \begin{example}[Abstract submission]
% \label{exm:abstract}
%
% The task of writing an abstract for a conference is never done.
% Let \TS{abstract} be a reference to a string.
%
% % \begin{TASK}
% %   let submit : String -> Task Unit =
% %     \text. delay 30 (sendTo "icfp19@sigplan.org" text)
% %   let write : Task Unit =
% %     edit "Abstract: ..." >>= \abstract. if 500 <= words abstract <= 700 then submit abstract else fail
% % \end{TASK}
%
% \begin{TASK}
%   let submit : Ref String -> Task String = \text. sendTo "icfp19@sigplan.org" text in
%   update abstract <|> delay 30 (submit abstract)
% \end{TASK}
%
% The \TS{delay} task is specified in \cref{exm:delay} in \cref{sec:language}.
%
% \end{example}

### 2019-02-27 sections/language.tex
use cleveref consistently
push fixmes in margin for last proofread

\label{sub:pretasks}
The grammar in \cref{fig:task-grammar} specifies the syntactic category of \emph{pretasks}.
Pretasks are tasks that have unevaluated subexpressions with respect to the host language.
Who expressions are evaluated will be discussed in Section~\ref{sec:evaluation}.
Each pretask will be discussed in more detail in the following subsections.
We use open symbols ($\Edit, \Enter, \Next, \Xor$) for tasks that require user input, and closed symbols ($\Update, \Then, \Or$) for tasks that can be evaluated without user input.

@@

Typing rules are of the form $\RelationT$, which should be read as \enquote{in environment $\Gamma$ and store typing $\Sigma$, expression $e$ has type $\tau$}.
Typing rules for expressions in the host language are presented in the appendix.
The typing rules for pretasks are given in \cref{fig:typing-rules}.

\fixme{Say someting about why we need beta, and use beta in the acutal typing rule}
\fixme{Alejandro: Is there anything surprising? Say what}
\begin{figure}[h]
  \small
  \begin{gather*}
    \boxed{\RelationT} \Break
    \userule{T-Edit} \Quad
    \userule{T-Enter} \Quad
    \userule{T-Update} \Quad
    \userule{T-Fail} \Break
    \userule{T-Then} \Quad
    \userule{T-Next} \Break
    \userule{T-And} \Quad
    \userule{T-Or} \Quad
    \userule{T-Xor}
  \end{gather*}
  \caption{Typing rules} \label{fig:typing-rules}
\end{figure}

@@

External choice ($\Xor$) is different in this regard.
An external choice requires users to pick a branch before continuing with it.
This means users cannot work on the branches of an external choice before picking one. \fixme{Alejandro: Is this actually implied?}

@@

% \subsection{Appointment}
%
% \fixme{write this subsection}

% It is essential for a workflow system to support multiple collaborators working
% together. Adding this funcitonaltiy to the \TOPHAT calculus is very
% straightforward. We identify users by means of a name $u$ which we can then use
% to label appointed tasks. The sole purpose of this labelling is to be able to
% tell which inputs are expected from what users. Other than that, it has no
% influence on the semantics. For space reasons, we have have excluded multi-user
% support from this paper.
%
% Mss moeten we het schrijven vanuit het oogpunt van labeling: je kunt labels
% toevoegen aan knoppen, als uitleg bij taken en om gebruikers toe te kennen aan
% taken. Vervolgens zou je iets met die gebruikers kunnen doen: alleen taken voor
% hun weergeven in de UI, alleen inputs voor hun genereren met I etc. Past dat?

@@

% For now, we only concentrate on appointing tasks to users,
% which is common in workflow modelling.
%
%
%
% \paragraph{Appoint $(u \At e)$}
%
% Users are identified by a name $u$ which is a string.
% They can be appointed tasks by means of the $\At$ combinator.
% The sole purpose of this labelling is to be able to tell users which inputs are expected of them.
% In \cref{sec:handling}, a function is described to calculate which users can take what actions.
% Other than this function, the appointment of tasks has no influence on the semantics.
% All semantic rules in \cref{sec:semantics} simply continue with their operations as if the label was not there.
% \todo{we still have to fix the type of $\Inputs$ and state the type of $u$}

>>>>

\label{sub:pretasks}
The grammar in \cref{fig:task-grammar} specifies the syntactic category of \emph{pretasks}.
Pretasks are tasks that have unevaluated subexpressions with respect to the host language.
Who expressions are evaluated will be discussed in \cref{sec:evaluation}.
Each pretask will be discussed in more detail in the following subsections.
We use open symbols ($\Edit, \Enter, \Next, \Xor$) for tasks that require user input, and closed symbols ($\Update, \Then, \Or$) for tasks that can be evaluated without user input.

@@

Typing rules are of the form $\RelationT$, which should be read as \enquote{in environment $\Gamma$ and store typing $\Sigma$, expression $e$ has type $\tau$}.
Typing rules for expressions in the host language are presented in the appendix.
The typing rules for pretasks are given in \cref{fig:typing-rules}.
\todo{Say someting about why we need beta, and use beta in the acutal typing rule}
\todo{Alejandro: Is there anything surprising? Say what}

\begin{figure}[h]
  \small
  \begin{gather*}
    \boxed{\RelationT} \Break
    \userule{T-Edit} \Quad
    \userule{T-Enter} \Quad
    \userule{T-Update} \Quad
    \userule{T-Fail} \Break
    \userule{T-Then} \Quad
    \userule{T-Next} \Break
    \userule{T-And} \Quad
    \userule{T-Or} \Quad
    \userule{T-Xor}
  \end{gather*}
  \caption{Typing rules} \label{fig:typing-rules}
\end{figure}

@@

External choice ($\Xor$) is different in this regard.
An external choice requires users to pick a branch before continuing with it.
This means users cannot work on the branches of an external choice before picking one. \todo{Alejandro: Is this actually implied?}

@@

% \subsection{Appointment}
%
%
% It is essential for a workflow system to support multiple collaborators working
% together. Adding this funcitonaltiy to the \TOPHAT calculus is very
% straightforward. We identify users by means of a name $u$ which we can then use
% to label appointed tasks. The sole purpose of this labelling is to be able to
% tell which inputs are expected from what users. Other than that, it has no
% influence on the semantics. For space reasons, we have have excluded multi-user
% support from this paper.
%
% Mss moeten we het schrijven vanuit het oogpunt van labeling: je kunt labels
% toevoegen aan knoppen, als uitleg bij taken en om gebruikers toe te kennen aan
% taken. Vervolgens zou je iets met die gebruikers kunnen doen: alleen taken voor
% hun weergeven in de UI, alleen inputs voor hun genereren met I etc. Past dat?

@@

% For now, we only concentrate on appointing tasks to users,
% which is common in workflow modelling.
%
%
%
% \paragraph{Appoint $(u \At e)$}
%
% Users are identified by a name $u$ which is a string.
% They can be appointed tasks by means of the $\At$ combinator.
% The sole purpose of this labelling is to be able to tell users which inputs are expected of them.
% In \cref{sec:handling}, a function is described to calculate which users can take what actions.
% Other than this function, the appointment of tasks has no influence on the semantics.
% All semantic rules in \cref{sec:semantics} simply continue with their operations as if the label was not there.

### 2019-02-27 sections/semantics.tex
some fixmes
F and S in inputs
use correct font for F and S in semantics
push fixmes in margin for last proofread

In this section we formalise the semantics of the language constructs described in \cref{sec:language}.
We organise this by following the structure of the language.

@@

\begin{figure}[h]
  \small
  \usemacro{G-Values-Compact}
  \caption{Value grammar} \label{fig:value-grammar}
\end{figure}

The rules to evaluate expressions $e$ do not differ from standard work, except for the task constructs.
The evaluation rules for tasks can be deduced from the value grammar.
They are given in the appendix.

@@

\fixme{Add $\UserInterface$ to appendix and refer.}
\TOPHAT is designed such that a user interface can be generated from a task's syntax tree.
A possible graphical user interface is shown in \cref{fig:flight-booking}, where tasks are rendered as \HTML pages.
Editors are rendered as input fields,
external choices are represented by two buttons,
and parallel tasks are rendered side by side.
Steps only show the interface of their left hand side.
In case of an external step they are accompanied by a button.
When the guard condition of a step is not fulfilled, the button is disabled.

@@

The split between striding and normalisation is due to mutable references.
Consider the following example, where $s=\set{l\mapsto\False}$.
\begin{TASK}
  (update l >>= \x:Bool. if x then e else fail) <&> (l := True; edit <<>>)
\end{TASK}
\refrule{S-And} reduces this expression in one step to
\begin{TASK}
  (update l >>= \x:Bool. if x then e else fail) <&> (edit <<>>)
\end{TASK}
with $s'=\set{l\mapsto\True}$.
This expression is not normalised, because the left task can take a step.
The issue here lies in the fact that the right task updates $l$.
To overcome this problem, the \refrule{N-Done} and \refrule{N-Repeat} rules ensure that striding is applied until the state $s$ becomes stable and no further normalisation can take place.

@@

Inputs $i$ are formed according to the grammar in \cref{fig:input-grammar}.
%
There is a function $\Inputs$ which calculates the possible input events a given task expects.
It takes a normalised task and a state and returns a set of inputs that can be handled.
The definition of this function is listed in \cref{fig:observation-input}.

\begin{figure}[h]
  \small
  \usemacro{G-Inputs-Compact}
  \caption{Input grammar} \label{fig:input-grammar}
\end{figure}

\begin{figure}[h]
  \small
  \usemacro{O-Inputs}
  \caption{Inputs} \label{fig:observation-input}
\end{figure}

>>>>

\todo{Alejandro: In the formalization you show quite some figures. I think that for
each figure you need to be more explicit in the text about what is
unsurprising and what is new or should get the attention of the reader.}
In this section we formalise the semantics of the language constructs described in \cref{sec:language}.
We organise this by following the structure of the language.

@@

\begin{figure}[h]
  \small
  \usemacro{G-Values-Compact}
  \caption{Value grammar} \label{fig:value-grammar}
\end{figure}
\todo{Alejandro over tasks: I was not expecting these to be values...}
The rules to evaluate expressions $e$ do not differ from standard work, except for the task constructs.
The evaluation rules for tasks can be deduced from the value grammar.
They are given in the appendix. \todo{Alejandro: What is interesting about them?}

@@

%FIXME: Add $\UserInterface$ to appendix and refer.
\TOPHAT is designed such that a user interface can be generated from a task's syntax tree.
A possible graphical user interface is shown in \cref{fig:flight-booking}, where tasks are rendered as \HTML pages.
Editors are rendered as input fields,
external choices are represented by two buttons,
and parallel tasks are rendered side by side.
Steps only show the interface of their left hand side.
In case of an external step they are accompanied by a button.
When the guard condition of a step is not fulfilled, the button is disabled.

@@

The split between striding and normalisation is due to mutable references.
Consider the following example, where $s=\set{l\mapsto\False}$.
\begin{TASK}
  (update l >>= \x:Bool. if x then e else fail) <&> (l := True; edit <<>>)
\end{TASK}
\refrule{S-And} reduces this expression in one step to
\begin{TASK}
  (update l >>= \x:Bool. if x then e else fail) <&> (edit <<>>)
\end{TASK}
with $s'=\set{l\mapsto\True}$.
This expression is not normalised, because the left task can take a step.
The issue here lies in the fact that the right task updates $l$.
To overcome this problem, the \refrule{N-Done} and \refrule{N-Repeat} rules ensure that striding is applied until the state $s$ becomes stable and no further normalisation can take place.
\todo{The example is not 100\% clear to me}

@@

Inputs $i$ are formed according to the grammar in \cref{fig:input-grammar}.
$\First$ and $\Second$ in an input encode the path to the task for which the input is designated.
%
There is a function $\Inputs$ which calculates the possible input events a given task expects.
It takes a normalised task and a state and returns a set of inputs that can be handled.
The definition of this function is listed in \cref{fig:observation-input}.

\begin{figure}[h]
  \small
  \usemacro{G-Inputs-Compact}
  \caption{Input grammar} \label{fig:input-grammar}
\end{figure}
\begin{figure}[h]
  \small
  \usemacro{O-Inputs}
  \caption{Inputs} \label{fig:observation-input}
\end{figure}
